Numerical simulation of the Cordilleran ice sheet \\ through the last glacial cycle

### 2014-07-07 kappa.tex
Various style corrections.

As ice flows and glide on its bed, it transports rock debris and erode the
landscape, thence leaving geomorphological traces of its former presence
(\cref{fig:glaciation-indicators}). In many parts of the world, montane people
and early explorers have since long learned to read these traces, and to
understand that glaciers had once been more extensive than today. With the
advent of the glacial theory, birthed the idea that, under colder temperatures
than today, expansive ice sheets had once covered much of Europe and North
America \citep{Agassiz.1840}.

@@

More recently, palaeo-climate records extracted from deep-sea sediments have
provided a much more detailed picture of the Earth environmental history
\citep[e.g.,][]{Emiliani.1955, Shackleton.Opdyke.1973}, thus indicating neither
one, nor four, but rather tens of glacial cycles. During the last million
years, glacial and interglacial cycles have succeeded each other with a
periodicity 100\,kyr \citep{Hays.etal.1976}. In the present thesis, we focus on
the most recent of these cycles: the last glacial cycle
(\cref{fig:paleo-timeseries})

% ice volume 26.07 + 2.93 = 29.00
% meters sea-level 58.3 + 0.70 = 59.0
Together, the Greenland and Antarctic ice sheets currently host most glacial
ice on the planet, corresponding to 59~meters of potential sea-level rise
\citep{Bamber.etal.2001, Fretwell.etal.2013}. At the apogee of the last glacial
cycle, the last glacial maximum, sea-level was 120 to 135\,m lower than today
\citep{Clark.Mix.2002}, implying an additional, double amount of ice stored in
ice sheets and glaciers on land. Systematic mapping and compilation of glacial
landforms from all continents permitted reconstructions of this former ice
cover \citep{Ehlers.Gibbard.2007}, indicating that ice volume
was repartitioned between more extensive Greenland and Antarctic ice sheets,
the Scandinavian and Barents-Kara ice sheets in the Eurasian arctic, and the
Innuitian, Laurentide and Cordilleran ice sheets over North America
(\cref{fig:paleo-glaciation}).

@@

% ----------------------------------------------------------------------
\subsection{Geographic setting}
% ----------------------------------------------------------------------

% fig:locmap
\begin{figure}
  \centering
  \makebox[0pt]{\includegraphics{map-northamerica}}
  \caption{Relief map of northern North America showing a reconstruction of the
           areas covered by the Cordilleran (CIS), Laurentide (LIS), Innuitian
           (IIS) and Greenland (GIS) ice sheets during the last glacial maximum
           \citep[21.4 to 16.8\,cal\,\chem{^{14}C}\,kyr\,BP][]{Dyke.2004}.
           The rectangular box delimits the model domain used throughout this
           thesis and on which most figures were drawn. The background
           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
           Data \citep{Patterson.Kelso.2014}.
           \todo{Mark palaeo-ice sheets on the map.}}
  \label{fig:locmap}
\end{figure}

\begin{figure}
  \centering
  \makebox[0pt]{\includegraphics{map-cordillera}}
  \caption{North-oriented topographic map of the northern American Cordillera.
           Cumulative last glacial maximum ice cover from \citet{Dyke.2004}.
           The background map consists of ETOPO1 \citep{Amante.Eakins.2009}
           and Natural Earth Data \citep{Patterson.Kelso.2014}.
           \todo{Mark the main reliefs. Add model domain?}}
  \label{fig:locmap}
\end{figure}

@@

@@

>>>>

As ice flows and glides on its bed, it transports rock debris and erode the
landscape, thence leaving geomorphological traces of its former presence
(\cref{fig:glaciation-indicators}). In many parts of the world, montane people
and early explorers have since long learned to read these traces, and to
understand that glaciers had once been more extensive than today. With the
advent of the glacial theory, birthed the idea that, under colder temperatures
than today, expansive \emph{ice sheets} had once covered much of Europe and
North America \citep{Agassiz.1840}.

@@

More recently, palaeo-climate records extracted from deep-sea sediments have
provided a much more detailed picture of the Earth environmental history
\citep[e.g.,][]{Emiliani.1955, Shackleton.Opdyke.1973}, thus indicating neither
one, nor four, but rather tens of glacial cycles. During the last million
years, glacial and interglacial cycles have succeeded each other with a
100\,kyr periodicity \citep{Hays.etal.1976}. In the present thesis, we focus on
the most recent of these cycles: \emph{the last glacial cycle}
(\cref{fig:paleo-timeseries})

% ice volume 26.07 + 2.93 = 29.00
% meters sea-level 58.3 + 0.70 = 59.0
Together, the Greenland and Antarctic ice sheets currently host most glacial
ice on the planet, corresponding to 59~meters of potential sea-level rise
\citep{Bamber.etal.2001, Fretwell.etal.2013}. At the apogee of the last glacial
cycle, the last glacial maximum, sea-level was 120 to 135\,m lower than today
\citep{Clark.Mix.2002}, implying an additional, double amount of ice stored in
ice sheets and glaciers on land. Systematic mapping and compilation of glacial
landforms from all continents permitted reconstructions of this former ice
cover \citep{Ehlers.Gibbard.2007}, indicating that ice volume
was repartitioned between more extensive Greenland and Antarctic ice sheets,
the Scandinavian and Barents-Kara ice sheets in the Eurasian arctic, and the
Innuitian, Laurentide and \emph{Cordilleran} ice sheets over North America
(\cref{fig:paleo-glaciation}).

@@

During the last glacial cycle, much of North America has been covered by ice.
At the last glacial maximum, apex of glaciation, three major ice sheets had
collided to form a continuous blanket of ice that spanned from Alaska to
Newfoundland, connecting with the Greenland ice sheet in the Canadian Arctic,
and extending as far south as Chicago and the Great Lakes
\cref{fig:map-northamerica}. The Laurentide ice
sheet, largest of the three, was centred on the modern Hudson Bay,
and fully covered the Canadian Prairies, Eastern Canada, and the Great Lakes
Basin. The Innuitian ice sheet, much smaller, covered the northernmost Canadian
archipelago. Finally, the Cordilleran ice sheet capped the mountain ranges of
the Western Cordillera, which forms our region of focus in the present thesis.

\begin{figure}
  \centering
  \makebox[0pt]{\includegraphics{map-northamerica}}
  \caption{Relief map of northern North America showing a reconstruction of the
           areas covered by the Cordilleran (CIS), Laurentide (LIS), Innuitian
           (IIS) and Greenland (GIS) ice sheets during the last glacial maximum
           \citep[21.4 to 16.8\,cal\,\chem{^{14}C}\,kyr\,BP][]{Dyke.2004}.
           The rectangular box delimits the model domain used throughout this
           thesis and on which most figures were drawn. The background
           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
           Data \citep{Patterson.Kelso.2014}.
           \todo{Mark palaeo-ice sheets on the map.}}
  \label{fig:map-northamerica}
\end{figure}

% ----------------------------------------------------------------------
\subsection{Topographic setting}
% ----------------------------------------------------------------------

In North American, the Western Cordillera forms a broad mountain system with an
average east-west width of about 1000\,km. Its topography is a complex
arrangement of mountain ranges and interior basins. It was formed during
multiple accretion phases associated with the closure of several back-arc
oceanic trenches over a 200\,Ma-long geological history \mref.

Most of the dominant reliefs are found along two major strips that run parallel
to each other along the general orientation of the Cordillera
\cref{fig:map-cordillera}. The western strip consists of the Pacific Coast
orogenic belt. Major ice-covered reliefs include,
from North to South, the Alaska Range, the Wrangell and Saint-Elias Mountains,
the Coast Mountains and the North Cascades. The eastern strip consists of the
Nevadan and Laramide belts, which are separated by the deep and strikingly
linear Rocky Mountain / Tintina Trench. It includes the Selwyn and
MacKenzie Mountains in the north, and the Rocky Mountains in the South. The
Brooks range, in northern Alaska, has been covered by ice, but its glaciers
were not connected to the main body of the Cordilleran ice sheet.

The Cordillera comports several intermontane basins that impacted patterns
of ice sheet flow. The two major depressions are the Interior Plateau of
central British Columbia, and the Liard Lowland. Between these two basins is
found a mildly elevated, but extensive and deeply incised mountain group, known
as the Skeena Mountains.

The ubiquitously mountainous topography of the region presents a serious
obstacle to reconstructing past dynamics of the Cordilleran ice sheet. Because
of the irregular arrangement of valleys and ridges that formed the ice sheet's
floor, ice moved along tortuous and intertwined pathways, sometimes following
the major topographical troughs, and sometimes flowing across them \mref. In
terms of
numerical glacier modelling, it implies complex patterns of flow and stress
within the ice, and requires high-resolution, and thus, large computational
grids.

\begin{figure}
  \centering
  \makebox[0pt]{\includegraphics{map-cordillera}}
  \caption{North-oriented topographic map of the northern American Cordillera.
           Cumulative last glacial maximum ice cover from \citet{Dyke.2004}.
           The background map consists of ETOPO1 \citep{Amante.Eakins.2009}
           and Natural Earth Data \citep{Patterson.Kelso.2014}.
           \todo{Mark the main reliefs. Add model domain?}}
  \label{fig:map-cordillera}
\end{figure}

@@

The topographic complexity of the North American Cordillera is associated with
distinctive patterns of temperature and precipitation, whose effect is often
clearly visible in the landscape (\cref{fig:photo-fraser-valley}). Due to the
large latitudinal extent of the region, mean annual temperatures vary
significantly from north to south (\cref{fig:plot-atm}a). About half of the
model domain has mean annual temperature below freezing. However, seasonality,
here computed as the temperature difference between the warmest and the coldest
months, varies between around 10\,\degC\ along most of the Pacific Coast to
above 40\,\degC\ in the polar interior (\cref{fig:plot-atm}b). There exist a
partial correlation between cold regions and those that have a large
seasonality, so that many regions where mean annual temperature is well below
freezing point do experience warm summers.

Precipitation rates are highest on the western slopes of the Pacific Coast
ranges and decline quickly on the other side of the mountains
(\cref{fig:plot-atm}c). In terms of ice sheet mass balance, this contrast is
even reinforced by the difference in timing of the precipitation peak through
the year (\cref{fig:plot-atm}d). Coastal regions experience most precipitation
during winter and fall, thus as snow. On the contrary, many inland regions
experience dry winters, and most of the annual precipitation consequently falls
as rain during the ablation season. The climatic setting of the North American
Cordillera is described further in \CCLI, where its effects on the ice sheet
model response have been studied in detail.

@@

During the last glacial maximum, the Cordilleran ice sheet covered most of the
Canadian Cordillera, and some of the adjacent land in north-western contiguous
USA and Alaska (\cref{map-cordillera}). The southern margin is the most
extensively studied. It consists of different piedmont lobes that have formed
at the outlet of the main valleys. The Puget lobe was the westernmost and
largest lobe. It covered the Puget Lowland, a region that currently includes
the cities of Vancouver and Seattle. East of the Cascade Range, were found the
Okanagan and Purcell lobes. The latter once dammed the Clark Fork River to form
Glacial Lake Missoula \mref. As the ice sheet retreated, the lake repeatedly
and catastrophically drained to the Pacific Ocean \mref. These floods
contributed to shape a distinctive landscape known as the Channeled Scablands.

The western margin of the ice sheet covered part of the continental shelf. An
ice cap covering Vancouver Island was well-connected to main body of ice on the
continent. A significant part of the ice sheet thus drained into the Georgia
Strait and engulfed between Vancouver Island and the Olympic Peninsula before
it formed the Juan the Fuca marine ice lobe, calving into the Pacific Ocean.
Further to the North, the ice margin reached Queen Charlotte Islands, yet the
archipelago may not have been fully covered by ice
flowing from the land. Submarine troughs likely idicate the erosional action of
former ice streams, yet it is not clear whether they have formed during the
last glacial cycle or earlier.

The northern margin of the ice sheet is in several ways more complex than the
southern margin. Firstly, geomorphological evidence indicate multiple glacier
advances. During the last glacial cycle, at least two glaciations occured, and
in some places, the first glaciation was more extensive than the second.
Secondly, geomorphological evidence indicate that the extent of glaciers in the
North was relatively restricted, despite very cold temperatures at present.

Finally, the eastern margin of the ice sheet is the most unclear. Although it
is known that the Cordilleran ice sheet collided with its Laurentide neighbour,
the precise location of this junction is unknown, and little information exists
on its timing. More importantly, the nature of the interaction between the two
ice sheets is uncertain. It is not known for instance, if the junction lasted
long enough for the ice sheets to significantly alter their respective flow
dynamics. Because the present thesis focuses specifically on the Cordilleran
ice sheet, our largest assumption certainly consists of neglecting buttressing
effects from the Laurentide ice sheet.

Although the last glacial maximum extent of the Cordilleran ice sheet, apart of
its eastern margin, is relatively well known, the history of deglaciation is
heavily debated, while little information exists on earlier dynamics of ice
sheet.
